Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-03-14 manuscript/ms.tex
Started writing Results section

\begin{figure*}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{pipeline/pipeline2}
  \caption{{\bf Diagrammatic representation of the various stages of the Ensembler pipeline.}
  The number of viable models surviving each stage of the pipeline for are shown, either for all tyrosine kinases (\emph{All TKs}) or representative individual kinases (\emph{SRC} and \emph{ABL}).
  In addition, the typical timing on a cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to convey resources required per model and for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single hyperthread---parallel execution can reduce wall clock time nearly linearly.
  }
  \label{figpipeline}
\end{figure*}

@@

The {\tt ensembler gather\_templates} subcommand also provides methods for selecting template structures from either UniProt or the Protein Data Bank (PDB; \href{http://www.rcsb.org/pdb}), specified by the {\tt -{}-gather\_from} flag.

@@

Selection of templates from the PDB simply requires passing a list of PDB IDs as a comma-separated string, e.g. {\tt -{}-query 2H8H,1Y57}.
% {\color{red}[JDC: Who specifies the list of PDB IDs?  How are they specified?  Can this be built automatically, or does the user need to specify these?]} DLP: these are specified by the user as a comma-separated list, as is now described above.
Specific PDB chain IDs can optionally also be selected via the {\tt -{}-chainids} flag.
The program retrieves structures from the PDB server, as well as associated data from the SIFTS service (\href{http://www.ebi.ac.uk/pdbe/docs/sifts/}{www.ebi.ac.uk/pdbe/docs/sifts}) (CITE: Velankar Nucleic Acids Res 2013), which provides residue-level mappings between PDB and UniProt entries.
The SIFTS data is used to extract template sequences, retaining only residues which are resolved and match the equivalent residue in the UniProt sequence---non-wildtype residues are thus removed from the template structures.
Furthermore, PDB chains with less than a given percentage of resolved residues (default: 70\%) are filtered out.
Sequences are stored in a FASTA file, with identifiers of the form {\tt [UniProt mnemonic]\_D[UniProt domain index]\_[PDB ID]\_[PDB chain ID]}, e.g. {\tt SRC\_HUMAN\_D0\_2H8H\_A}.
Template structures with residues matching the sequence data are then extracted and stored as PDB-format coordinate files.

@@

Unresolved template loops can optionally be remodeled with a kinematic closure algorithm {\color{red}[CITE]}, which is provided via the loopmodel tool of the Rosetta software suite (CITE: Rosetta and/or loopmodel).
Because fewer loops need to be built during the subsequent model-building stage, prebuilding template loops tends to provide higher-quality models following the subsequent modeling process.

@@

% ? The API is highly flexible, allowing control over many important parameters. Appropriate choices have been set as default values.

@@

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{seqid_dist/seqid_dist}
    \caption{{\bf Sequence identity distribution for human TK models.}
    Distribution of sequence identities for all 373,513 models generated for the human tyrosine kinases.
    {\color{blue}[DLP: should I mention the use of KDE smoothing?]}
    Sequence identities are calculated from pairwise target-template alignments.
    The cumulative distribution function is shown by the dashed line.
    }
  \label{figseqid_dist}
\end{figure}

@@

\begin{figure}
  \centering
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/distances}
      \caption{Src}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/abl/distances}
      \caption{Abl}
    \end{subfigure}
    \caption{{\bf Residue pair distances for Src and Abl.}
    % ...
  }
  \label{figaa_pair_distances_ref_structures}
\end{figure}

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/src_ref_structures}
    \caption{{\bf Reference structures for Src.}
    % ...
  }
  \label{figsrc_ref_structures}
\end{figure}

@@

>>>>

\begin{figure*}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{pipeline/pipeline2}
  \caption{{\bf Diagrammatic representation of the various stages of the Ensembler pipeline.}
  The number of viable models surviving each stage of the pipeline are shown, either for all tyrosine kinases (\emph{All TKs}) or representative individual kinases (\emph{SRC} and \emph{ABL}).
  In addition, the typical timing on a cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to exemplify the resources required per model and for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single hyperthread---parallel execution can reduce wall clock time nearly linearly.
  }
  \label{figpipeline}
\end{figure*}

@@

The {\tt ensembler gather\_templates} subcommand also provides methods for selecting template structures from either UniProt or the Protein Data Bank (PDB; \href{http://www.rcsb.org/pdb}), specified by the {\tt -{}-gather\_from} flag.
Both methods select templates at the level of PDB chains---a PDB structure containing multiple chains with identical sequence spans (e.g. for crystal unit cells with multiple asymmetric units) would thus give rise to multiple template structures.

@@

Selection of templates from the PDB simply requires passing a list of PDB IDs as a comma-separated string, e.g. {\tt -{}-query 2H8H,1Y57}.
% {\color{red}[JDC: Who specifies the list of PDB IDs?  How are they specified?  Can this be built automatically, or does the user need to specify these?]} DLP: these are specified by the user as a comma-separated list, as is now described above.
Specific PDB chain IDs can optionally also be selected via the {\tt -{}-chainids} flag.
The program retrieves structures from the PDB server, as well as associated data from the SIFTS service (\href{http://www.ebi.ac.uk/pdbe/docs/sifts/}{www.ebi.ac.uk/pdbe/docs/sifts}) (CITE: Velankar Nucleic Acids Res 2013), which provides residue-level mappings between PDB and UniProt entries.
The SIFTS data is used to extract template sequences, retaining only residues which are resolved and match the equivalent residue in the UniProt sequence---non-wildtype residues are thus removed from the template structures.
Furthermore, PDB chains with less than a given percentage of resolved residues (default: 70\%) are filtered out.
Sequences are stored in a FASTA file, with identifiers of the form {\tt [UniProt mnemonic]\_D[UniProt domain index]\_[PDB ID]\_[PDB chain ID]}, e.g. {\tt SRC\_HUMAN\_D0\_2H8H\_A}.
Matching residues then extracted from the original coordinate files and stored as PDB-format coordinate files.

@@

Unresolved template residues can optionally be remodeled with the {\tt loopmodel} subcommand, which employs a kinematic closure algorithm {\color{red}[CITE]} provided via the {\tt loopmodel} tool of the Rosetta software suite (CITE: Rosetta and/or loopmodel).
Because fewer loops need to be built during the subsequent model-building stage, prebuilding template loops tends to provide higher-quality models after completion of the {\bf Ensembler} pipeline.
Loop remodeling may fail for a small proportion of templates due to spatial constraints imposed by the original structure; the subsequent modeling step thus automatically uses the remodeled version of a template if available, but otherwise falls back to using the non-remodeled version.
Furthermore, the Rosetta {\tt loopmodel} program will not model missing residues at the termini of a structure---such residues spans are modeled in the subsequent stage.

@@

% TODO? The API is highly flexible, allowing control over many important parameters. Appropriate choices have been set as default values.

% TODO? other tools for analyzing project etc.

@@

As a first application of {\bf Ensembler}, we have built models for all 90 human tyrosine kinase (TKs) domains listed in UniProt.
TKs (and protein kinases in general) play important roles in many cellular processes and are involved in a number of types of cancer.
For example, mutations of Src are associated with colon, breast and prostate cancer {\color{red}[CITE: Src cancer involvement]}, while a translocation between the TK Abl and the pseudokinase Bcr is closely associated with chronic myelogenous leukemia {\color{red}[CITE: Abl cancer involvement]}.
Protein kinase domains are thought to have multiple accessible metastable conformation states, with a single active conformation, and much effort is directed at developing kinase inhibitor drugs which bind to and stabilize inactive conformations [CITE: Lee and Craik Science 2009].
Kinases are thus a particularly interesting subject for study with MSM methods [CITE: recent kinase MSM papers], and this approach could benefit greatly from the ability to exploit the full body of available genomic and structural data within the kinase family, e.g. by generating large numbers of starting configurations to be used in highly parallel MD simulation.

We selected all available structures of protein kinase domains (of any species) as templates, for a total of 4433 (\num{398970} target-template pairs).
The templates were derived from 3028 individual PDB entries and encompassed 23 different species, with 3634 template structures from human kinase constructs.

The templates were first subjected to loop remodeling, using the {\tt loopmodel} subcommand.
Out of 3666 templates with one or more missing residues, 3134 were successfully remodeled.

Following loop remodeling, the {\bf Ensembler} pipeline was performed up to and including the implicit solvent MD refinement stage, which completed with \num{373513} surviving models.
In addition, the solvation stage was performed for two representative individual kinases (\emph{Src} and \emph{Abl}).
The number of models which survived each stage are shown in Fig. \ref{figpipeline}, indicating that the greatest attrition occurred during the modeling stage.
Fig. \ref{figpipeline} also indicates the typical timing achieved on a cluster for each stage.

The distribution of RMSDs of the final models (relative to the highest sequence identity model for a given target) is shown in Fig. \ref{figrmsddist}.
The distributions are stratified based on the sequence identity between target and template, indicating that higher sequence identity templates result in lower RMSD models.
The sequence identity stratifications were selected based on the sequence identity distribution (Fig. \ref{figseqiddist}), which suggests an intuitive division into three categories, with \num{307753} models in the 0-35\% sequence identity range, \num{69922} models in the 35-55\% range, and \num{4893} models in the 55-100\% range.

% superpositions etc.

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{seqid_dist/seqid_dist}
    \caption{{\bf Sequence identity distribution for human TK models.}
    Distribution of sequence identities for all 373,513 models generated for the human tyrosine kinases.
    {\color{blue}[DLP: should I mention the use of KDE smoothing?]}
    Sequence identities are calculated from pairwise target-template alignments.
    The cumulative distribution function is shown by the dashed line.
    }
  \label{figseqiddist}
\end{figure}

@@

\begin{figure}
  \centering
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/distances}
      \caption{Src}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/abl/distances}
      \caption{Abl}
    \end{subfigure}
    \caption{{\bf E310-R409 and K295-E310 distances for models of Src and Abl, colored by sequence identity.}
  }
  \label{figaa_pair_distances_ref_structures}
\end{figure}

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/src_ref_structures}
    % \begin{subfigure}[b]{0.45\textwidth}
    %   \includegraphics[width=1.0\textwidth, clip=true, trim=30em 0em 40em 0em]{residue_pair_distances/src/inactive}
    %   \caption{Inactive}
    % \end{subfigure}
    % \begin{subfigure}[b]{0.45\textwidth}
    %   \includegraphics[width=1.0\textwidth, clip=true, trim=30em 0em 27em 10em]{residue_pair_distances/src/active}
    %   \caption{Active}
    % \end{subfigure}
    \caption{{\bf Two structures of Src, indicating certain residues involved in activation.}
    In the inactive state, E310 forms a salt bridge with R409.
    During activation, the C-helix (green) moves and rotates, orienting E310 towards the ATP-binding site and allowing it to instead form a salt bridge with K295.
    This positions K295 in the appropriate position for catalysis.
  }
  \label{figsrc_ref_structures}
\end{figure}

@@

The latest release of {\bf Ensembler} can be installed via the {\tt conda} package manager for Python.
Instructions can be found at \url{https://github.com/choderalab/ensembler}.
This will install all dependencies except for Modeller and (optionally) Rosetta, which are not available through the {\tt conda} package manager, and thus must be installed separately by the user.
The latest source can be downloaded from the above GitHub repository, which also contains instructions for building and installing the code.

% TODO future directions
% TODO what's on the Ensembler 2 roadmap?
% TODO MCCE?
